---
title: "sarcoidosis scRNAseq processing"
author: "Titus Schluter"
date: last-modified
format: 
  html: 
    toc: true
    code-fold: show
    embed-resources: true
engine: knitr
editor: source
---

# Seurat

```{r setup}
# Loading packages.
library(KEGGREST)
library(fgsea)
library(reshape2)
library(msigdbr)
library(scDblFinder)
library(Seurat)
library(grDevices)
library(RColorBrewer)
library(ggrepel)
library(ggpmisc)
library(ggh4x)
library(patchwork)
library(Matrix)
library(DESeq2)
library(pheatmap)
library(ggpubr)
library(org.Hs.eg.db)
library(AnnotationHub)
library(sctransform)
library(ggbeeswarm)
library(ensembldb)
library(tidyverse)

setOldClass("ExpData")

# Setting working directory.
setwd("/mnt/data/2_Users/Titus_Schluter/sarcoidosis_rna/")

sarcoidosis_palette <- c("healthy" = "#e7e3d8",
                         "Stable persistant" = "#9fc5e8",
                         "Stable_persistant" = "#9fc5e8",
                         "stable" = "#9fc5e8",
                         "Progressive" = "#1d4c6e",
                         "sarcoidosis" = "#1d4c6e",
                         "progressive" = "#1d4c6e",
                         "low_non" = "#b2d8d8",
                         "responder" = "#006666"
                         )
```

## Set up annotationhub
```{r setup}
# Preparing annotation file with Ensembl and Entrez IDs.
ah <- AnnotationHub()

human_ens <- query(ah, c("Homo sapiens", "EnsDb"))

human_ens <- human_ens[["AH116860"]]

gene_annotation <- genes(human_ens, return.type = "data.frame") |> 
  select(gene_id, gene_name, entrezid, gene_biotype)

gene_annotation$entrezid <- map(gene_annotation$entrezid, 1) |> 
  unlist()

rm(ah, human_ens)
```

# Preprocessing
## Reading in count matrices and converting to seurat object (done)
```{r}
# Reading in count matrix. For some reason it is not working for RPMI_baseline sample, so we need to figure that out.
sarcoidosis <- sapply(list.files("per_sample_outs/"), function(sample){
  paste0("per_sample_outs/", sample, "/count/sample_filtered_feature_bc_matrix") |>
    Read10X()
  }, USE.NAMES = T)

sarcoidosis <- sapply(sarcoidosis, function(sample){CreateSeuratObject(sample)}, USE.NAMES = T)

sarcoidosis <- sapply(names(sarcoidosis), function(name){sarcoidosis[[name]]$sampleID <- name
sarcoidosis[[name]]})

sarcoidosis <- merge(x = sarcoidosis[[1]], y = sarcoidosis[2:length(sarcoidosis)])
```

## Quality control
```{r, eval = F}
# Calculating QC metric for proportion of mitochondrial reads per cell.
sarcoidosis_subset <- subset(sarcoidosis, sampleID == "A-01" |
                      sampleID == "A-02" |
                      sampleID == "HC_N_01" |
                      sampleID == "HC_N_02" |
                      sampleID == "Healthy_2_tube1-2" |
                      sampleID == "Healthy_4" |
                      sampleID == "HSJ-005" |
                      sampleID == "HSJ-032" |
                      sampleID == "HSJ-045" |
                      sampleID == "HSJ-073" |
                      sampleID == "HSJ-221" |
                      sampleID == "HSJ-222" |
                      sampleID == "HSJ-224" |
                      sampleID == "HSJ-226" |
                      sampleID == "HSJ-227" |
                      sampleID == "HSJ-228" |
                      sampleID == "HSJ-230" |
                      sampleID == "HSJ-231")



sarcoidosis_subset$mitoRatio <- PercentageFeatureSet(object = sarcoidosis_subset, pattern = "^MT-") / 100


VlnPlot(sarcoidosis_subset, features = c("mitoRatio"), group.by = "sampleID", pt.size = 0,y.max = 0.2)


# Updating metadata, adding cellID, sample metadata, and other metrics.
metadata <- sarcoidosis_subset@meta.data
metadata$cellID <- rownames(metadata)

metadata <- metadata |> 
  rename(nUMI = nCount_RNA,
         nGene = nFeature_RNA) |>
  mutate(log10GenesPerUMI = log10(nGene) / log10(nUMI),
         log10UMI = log10(nUMI),
         log10Gene = log10(nGene))

sarcoidosis_subset@meta.data <- metadata

# Plotting default Seurat QC plots. Data appears generally high quality, with many genes/RNA molecules detected per cell and most cells having low mitochondrial gene expression.
VlnPlot(sarcoidosis_subset, features = c("nGene", "mitoRatio", "log10UMI"), pt.size = 0, group.by = "sampleID")

FeatureScatter(sarcoidosis_subset, feature1 = "log10UMI", feature2 = "log10Gene", plot.cor = F) +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)),
                     labels = scales::comma) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0))) +
  NoLegend()

FeatureScatter(sarcoidosis_subset, feature1 = "log10UMI", feature2 = "log10GenesPerUMI", plot.cor = F) +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)),
                     labels = scales::comma) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)),
                     labels = scales::comma) +
  NoLegend() +
  stat_function(fun = function(log10UMI) -0.085 * log10UMI + 1.18) +
  geom_vline(xintercept = 3)

sarcoidosis_filtered <- sarcoidosis_subset |> 
  subset(log10GenesPerUMI > 0.75) |> 
  subset(mitoRatio <= 0.07) |> 
  subset(nGene >= 1000) |> 
  subset(log10GenesPerUMI > -0.085 * log10UMI + 1.18)


ncol(sarcoidosis_subset) - ncol(sarcoidosis_filtered)
(ncol(sarcoidosis_subset) - ncol(sarcoidosis_filtered)) / ncol(sarcoidosis_subset)


VlnPlot(sarcoidosis_filtered, features = c("nGene", "mitoRatio", "log10UMI"), pt.size = 0, group.by = "orig.ident") &
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)),
                     labels = scales::comma) &
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank())
```

```{r, eval = F}
# Determining which cells might be doublets. Assuming 3% of cells as per Parse FAQ.
# Merging the layers into a single object.
sarcoidosis_filtered[["RNA"]] <- JoinLayers(sarcoidosis_filtered[["RNA"]])

filtered_sce <- as.SingleCellExperiment(sarcoidosis_filtered)
filtered_sce <- scDblFinder(filtered_sce, dbr = 0.05)

sarcoidosis_filtered$scDblFinder.score <- filtered_sce$scDblFinder.score

metadata <- sarcoidosis_filtered@meta.data

metadata |>
  mutate(Threshold = scDblFinder.score > 0.25) |>
  ggplot(aes(x = scDblFinder.score, fill = Threshold)) +
  geom_histogram() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_bw()

sarcoidosis_filtered <- sarcoidosis_filtered |> 
  subset(scDblFinder.score < 0.25)

# Analyzing clean metadata: Removed  cells.
metadata <- sarcoidosis_filtered@meta.data

# Plotting number of cells per sample.
sarcoidosis_filtered@meta.data |>
  ggplot(aes(x = sampleID)) + 
	geom_bar() +
	labs(y = "Number of cells") +
  scale_y_continuous(expand = expansion(c(0, 0.05)), label = scales::comma) +
  theme_bw() +
	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
	      axis.title.x = element_blank(),
	      panel.grid.major.x = element_blank(),
	      legend.title = element_blank())
```

## Clustering/integration
```{r}
sarcoidosis_filtered[["RNA"]] <- split(sarcoidosis_filtered[["RNA"]], f = sarcoidosis_filtered$sampleID)

sarcoidosis_integrated <- sarcoidosis_filtered |> 
  NormalizeData() |> 
  FindVariableFeatures(nFeatures = 2000) |>
  ScaleData() |> 
  RunPCA() |> 
  FindNeighbors(dims = 1:40, reduction = "pca") |> 
  FindClusters(resolution = 1, cluster.name = "clusters.1.unintegrated")



sarcoidosis_integrated <- RunUMAP(sarcoidosis_integrated,
                                  dims = 1:40,
                                  reduction = "pca",
                                  reduction.name = "umap.unintegrated")

sarcoidosis_integrated <- IntegrateLayers(sarcoidosis_integrated,
                                          method = HarmonyIntegration,
                                          orig.reduction = "pca",
                                          new.reduction = "harmony")

sarcoidosis_integrated <- FindNeighbors(sarcoidosis_integrated,
                                        dims = 1:40,
                                        reduction = "harmony") |> 
                          FindClusters(resolution = 1,
                                       cluster.name = "clusters.1.harmony")

sarcoidosis_integrated <- RunUMAP(sarcoidosis_integrated,
                      dims = 1:40,
                      reduction = "harmony",
                      reduction.name = "umap.harmony")

saveRDS(sarcoidosis_integrated, "sarcoidosis_integrated.rds")

#sarcoidosis_integrated <- readRDS("sarcoidosis_integrated.rds")
```

## Cluster Annotation 

### Cluster identification
```{r}
# Labeling our clusters based on Azimuth labels.
Idents(sarcoidosis_integrated) <- sarcoidosis_integrated$clusters.1.harmony

tmp <- tibble(Var1 = "",
              Freq = 0,
              cluster_id = 0) |> 
  slice(-1)

for(i in unique(sarcoidosis_integrated@meta.data$clusters.1.harmony)) {
  tmp1 <- table(sarcoidosis_integrated@meta.data |> filter(clusters.1.harmony == i) |> pull(predicted.celltype.l2)) |> as.data.frame() |> mutate(cluster_id = i)
  
  tmp <- rbind(tmp, tmp1)
}

tmp <- tmp |> 
  select(Cluster_ID = cluster_id, predicted.celltype.l2 = Var1, Freq) |> 
  mutate(Cluster_ID = as.integer(Cluster_ID)) |> 
  arrange(Cluster_ID, desc(Freq))

tmp2 <- tmp |> 
  group_by(Cluster_ID) |> 
  summarize(N = sum(Freq))

tmp3 <- tmp |> 
  left_join(tmp2) |> 
  mutate(Percent = (Freq / N) * 100) |> 
  filter(Percent > 20)

tmp4 <- tmp3 |> 
  select(Cluster_ID, predicted.celltype.l2) |> 
  group_by(Cluster_ID) |> 
  summarise(cluster_names1 = paste(predicted.celltype.l2, collapse = "/")) |> 
  mutate(cluster_names1 = make.unique(cluster_names1, sep = " "),
         cluster_names1 = factor(cluster_names1),
         Cluster_ID = factor(Cluster_ID))

sarcoidosis_integrated@meta.data <- sarcoidosis_integrated@meta.data |> 
  #select(-cluster_names1) |> 
  left_join(tmp4, by = c("clusters.1.harmony" = "Cluster_ID")) |> 
  as.data.frame()

rownames(sarcoidosis_integrated@meta.data) <- sarcoidosis_integrated@meta.data$cellID
```

```{r}
# Plotting canonical markers for PBMC cell types in flow. Based on antibodies-online CD marker panel table, OMIP-69 flow panel, and 50 marker panel. Found genes using genecards. Some flow markers are combinations of different proteins, so can't just use a single gene.
# CD3/CD2 = T cells (CD3G/CD3D/CD3E/CD2)
# CD4 = T helper cells (CD4)
# CD8 = cytotoxic T cells (CD8A/CD8B)
# CCR7 = Naive T (CCR7)
# CD25/FOXP3 = Treg (IL2RA/FOXP3)
# TCRgd = gdT (TRG and TRD)
# CXCR5 = Tfh (also sometimes CXCR3 and CCR6)
# IL17A = Th17
# CD69 = Activated T cells
# CD161 = MAIT (KLRB1)
# CD56 = NK cells (NCAM1)
# CD19/CD20 = B cells (CD19/MS4A1)
# IGD/IGM = naive B cells (IGHD/IGHM)
# CD27 = plasmablast
# HLA-DR = Antigen presenting cell (HLA-DRA)
# CD14 = Classical monocytes (CD14)
# CD16 = Non-classical monocytes and mature natural killers (FCGR3A) 
# CD11C = dendritic cells (ITGAX)
# CD1c = cDC1 (CD1C)
# CD123 = plasmacytoid dendritic cells (IL3RA)
# CD34 = HSPC (CD34)
# CD33 + ITGAM (CD11B) = macrophage
```

```{r}
sarcoidosis_integrated@meta.data <- sarcoidosis_integrated@meta.data |> 
  mutate(cluster_names_manual = case_when(cluster_names1 == "NK" ~ "NK",
                                          cluster_names1 == "NK Proliferating/dnT" ~ "Proliferating",
                                          cluster_names1 == "MAIT" ~ "MAIT",
                                          cluster_names1 == "CD14 Mono/CD16 Mono" ~ "Mono_strange",
                                          cluster_names1 == "CD16 Mono" ~ "CD16_Mono",
                                          cluster_names1 == "HSPC" ~ "HSPC",
                                          cluster_names1 == "CD14 Mono" ~ "CD14_Mono_1",
                                          cluster_names1 == "CD14 Mono 1" ~ "CD14_Mono_2",
                                          cluster_names1 == "CD14 Mono 2" ~ "CD14_Mono_3",
                                          cluster_names1 == "CD14 Mono 3" ~ "CD14_Mono_Platelet",
                                          cluster_names1 == "pDC" ~ "pDC",
                                          cluster_names1 == "cDC2/CD14 Mono" ~ "DC",
                                          cluster_names1 == "NK/NK_CD56bright" ~ "early_NK",
                                          cluster_names1 == "B naive" ~ "B_naive",
                                          cluster_names1 == "B intermediate/B memory" ~ "B_memory",
                                          cluster_names1 == "Plasmablast" ~ "Plasmablast",
                                          cluster_names1 == "Treg/CD4 TCM" ~ "Treg",
                                          cluster_names1 == "CD8 Naive" ~ "CD8_naive",
                                          cluster_names1 == "CD4 TCM" ~ "CD4_memory",
                                          cluster_names1 == "CD8 TEM" ~ "CD8_memory",
                                          cluster_names1 == "CD8 TEM 1" ~ "CD8_memory",
                                          cluster_names1 == "NK/CD8 TEM 1" ~ "NKT strange",
                                          cluster_names1 == "NK/CD8 TEM" ~ "NKT_like",
                                          cluster_names1 == "CD4 TCM/CD4 Naive" ~ "CD4_naive",
                                          cluster_names1 == "CD4 TCM/CD4 Naive 1" ~ "CD4_naive",
                                    TRUE ~ cluster_names1
                                    
                                    )
         )

sarcoidosis_integrated@meta.data <- sarcoidosis_integrated@meta.data |> 
  mutate(cluster_names_pseudobulk = case_when(cluster_names1 == "NK" ~ "NK",
                                          cluster_names1 == "NK Proliferating/dnT" ~ "excl",
                                          cluster_names1 == "MAIT" ~ "MAIT",
                                          cluster_names1 == "CD14 Mono/CD16 Mono" ~ "excl",
                                          cluster_names1 == "CD16 Mono" ~ "CD16_Mono",
                                          cluster_names1 == "HSPC" ~ "excl",
                                          cluster_names1 == "CD14 Mono" ~ "CD14_Mono",
                                          cluster_names1 == "CD14 Mono 1" ~ "CD14_Mono",
                                          cluster_names1 == "CD14 Mono 2" ~ "CD14_Mono",
                                          cluster_names1 == "CD14 Mono 3" ~ "excl",
                                          cluster_names1 == "pDC" ~ "pDC",
                                          cluster_names1 == "cDC2/CD14 Mono" ~ "DC",
                                          cluster_names1 == "NK/NK_CD56bright" ~ "early_NK",
                                          cluster_names1 == "B naive" ~ "B_naive",
                                          cluster_names1 == "B intermediate/B memory" ~ "B_memory",
                                          cluster_names1 == "Plasmablast" ~ "excl",
                                          cluster_names1 == "Treg/CD4 TCM" ~ "Treg",
                                          cluster_names1 == "CD8 Naive" ~ "CD8_naive",
                                          cluster_names1 == "CD4 TCM" ~ "CD4_memory",
                                          cluster_names1 == "CD8 TEM" ~ "CD8_memory",
                                          cluster_names1 == "CD8 TEM 1" ~ "CD8_memory",
                                          cluster_names1 == "NK/CD8 TEM 1" ~ "excl",
                                          cluster_names1 == "NK/CD8 TEM" ~ "NKT_like",
                                          cluster_names1 == "CD4 TCM/CD4 Naive" ~ "CD4_naive",
                                          cluster_names1 == "CD4 TCM/CD4 Naive 1" ~ "CD4_naive",
                                    TRUE ~ cluster_names1
                                    
                                    )
         )

```

```{r}
Idents(sarcoidosis_integrated) <- sarcoidosis_integrated$cluster_names_pseudobulk
p1 <- DimPlot(object = sarcoidosis_integrated, 
              reduction = "umap.harmony", 
              label = TRUE,
              label.size = 3,
              raster = F) + NoLegend()

p2 <- DotPlot(sarcoidosis_integrated, features =  c("CD3E", "CD4", "CD8A", "CCR7", "CD69", "IL2RA", "FOXP3", "KLRB1", "NCAM1", "FCGR3A", "CD19", "MS4A1", "IGHD", "IGHM", "CD27", "CD33", "ITGAM", "CD14", "ITGAX", "CD1C", "IL3RA", "CD34", "IL7R", "MKI67", "CCR4", "CCL5", "ITGA2B"), cluster.idents = T, ) +
  scale_color_gradient2(low = "blue", mid = "grey90", high = "red") +
  scale_radius(limits = c(0, 100)) +
  RotatedAxis() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank())


p1 + p2

# ggsave("output/celltype_annotation.pdf", height = 8, width = 16)
```

### Pseudobulk: Create Pseudobulk DESeqDataSets
```{r}
pops <- sarcoidosis_integrated$cluster_names_pseudobulk |> unique()

library(DESeq2)
pseudobulk <- sapply(pops, function(pop){
bulk <- sarcoidosis_integrated |> 
  subset(state != "CGD") |> 
  subset(cluster_names_pseudobulk == pop) |>
AggregateExpression(group.by = "SampleID", return.seurat = F, normalization.method = NULL)

bulk <- bulk$RNA
colnames(bulk) <- str_replace_all(colnames(bulk), "-", "_")
colnames(bulk)[1:4] <- str_replace(colnames(bulk)[1:4], "g", "")

libsize <- bulk |>
  as.data.frame() |>
  summarise(across(everything(), ~ sum(.x))) |>
  t() |>
  as.data.frame() |>
  rename(libsize = V1) |> 
  rownames_to_column("SampleID")

sample_info_bulk <- sample_info |>
  mutate(state = factor(state, levels = c("healthy", "sarcoidosis"))) |> left_join(libsize)

bulk_anno <- data.frame(SampleID = colnames(bulk)) |> 
  left_join(sample_info_bulk)

dds <- DESeqDataSetFromMatrix(bulk, bulk_anno, ~state)

dds <- DESeq(dds)
return(dds)
})

write_rds(pseudobulk, "output/pseudobulk.rds")
```

### Pseudobulk: PW analysis
```{r}
library(fgsea)
library(msigdbr)

pops <- list.files("output/pseudobulk/")

hallmarks <- msigdbr(collection = "H")
go <- msigdbr(collection = "C5", subcollection = "GO:BP")
kegg <- msigdbr(collection = "C2", subcollection = "CP:KEGG_LEGACY")
reactome <- msigdbr(collection = "C2", subcollection = "CP:REACTOME")

allpws <- rbind(hallmarks, go, kegg, reactome)
allpws_gsea <- split(allpws, f = allpws$gs_name)
allpws_gsea <- sapply(allpws_gsea, function(pathway){
  pathway |> pull(gene_symbol)
})

gse_hm <- lapply(names(pseudobulk), function(pop){
  
gse_df <- pseudobulk[[pop]] |> 
  results(contrast = c("state", "sarcoidosis", "healthy")) |> 
  as.data.frame() |> 
  subset(baseMean > 50) |>
  arrange(-stat)

ranked_genes <- gse_df$stat
names(ranked_genes) <- rownames(gse_df)

#gse_list <- gse_list[!is.na(names(gse_list))]
set.seed(42)
gse_hm <- fgsea(pathways = allpws_gsea, stats = ranked_genes) |> 
  as.data.frame() |> 
  mutate(celltype = pop)

return(gse_hm)
}) |>
  bind_rows()

write_csv(gse_hm, "output/gsea_allpws.csv")
```

### Pseudobulk: Export/count DE genes and GO terms
```{r}
pops <- list.files("output/pseudobulk/")

all_fc <- lapply(pops, function(pop){
readRDS(paste0("output/pseudobulk/", pop, "/dds.rds")) |> 
    results(name = "state_sarcoidosis_vs_healthy") |> 
    as.data.frame() |> 
    mutate(direction = case_when(log2FoldChange > 0 ~ "up",
                               log2FoldChange < 0 ~ "down"),
           celltype = pop) |> 
    rownames_to_column("gene")
}) |> bind_rows()

write_csv(all_fc, "output/all_fc.csv")
```

# SCENIC+
## Create AnnData for SCENIC+
```{r}
sarcoidosis_sce <- as.SingleCellExperiment(sarcoidosis_integrated)

# install.packages("pak")
# pak::pak(c("scverse/anndataR"))

adata <- anndataR::as_AnnData(sarcoidosis_sce)

adata$write_h5ad("sarcoidosis_integrated.h5ad")
```

##Run GSEA on SCENIC+ regulons
```{r}
all_fc <- read_csv("output/all_fc.csv")

eregulons <- read_tsv("../sarcoidosis_atac/outs_scplus/eRegulon_direct.tsv")

eregulons_gsea <- eregulons |> 
  dplyr::select(Gene, eRegulon_name) |> 
  split(~eRegulon_name) |> 
  sapply(function(set){
  set$Gene
})

gse_hm_eregulons <- lapply(names(pseudobulk), function(pop){
  
gse_df <- all_fc |> 
  subset(celltype == pop) |> 
  subset(baseMean > 50) |>
  arrange(-stat)

ranked_genes <- gse_df$stat
names(ranked_genes) <- gse_df$gene

#gse_list <- gse_list[!is.na(names(gse_list))]

gse_hm <- fgsea(pathways = c(eregulons_gsea, eregulons_extended_gsea), stats = ranked_genes, eps = 0) |> 
  as.data.frame() |> 
  mutate(celltype = pop)

return(gse_hm)
}) |> bind_rows()
```

## Generate AUC for scRNAseq data
```{r}
library(AUCell)

sarcoidosis_integrated[["RNA"]]$data

cells_AUC <- AUCell_run(sarcoidosis_integrated[["RNA"]]$data, geneSets = eregulons_gsea)

cells_AUC_mtx <- getAUC(cells_AUC)

sarcoidosis_integrated[["AUC"]] <- CreateAssayObject(counts = cells_AUC_mtx)

write_rds(sarcoidosis_integrated, "sarcoidosis_integrated.rds")
```
