---
title: "Sarcoidosis_snATAC 2"
format: html
editor: source
---

# R setup
```{r}
library(ggpmisc)
library(msigdbr)
library(ggrepel)
library(ggh4x)
library(harmony)
library(JASPAR2020)
library(TFBSTools)
library(DESeq2)
library(motifmatchr)
library(BSgenome.Hsapiens.UCSC.hg38)
library(clusterProfiler)
library(ggpubr)
library(pheatmap)
library(ggbeeswarm)
library(AnnotationHub)
library(GenomicRanges)
library(Seurat)
library(Signac)
library(org.Hs.eg.db)
library(tidyverse)

setwd("/mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/")
```

```{r}
#sarc_atac_integrated <- readRDS("sarc_atac_integrated.rds")
#sarc_atac_pctopic <- readRDS("sarc_atac_pctopic.rds")
```

# Pseudobulk
### Using PyCisTopic consensus peaks for bulk analysis:
While it was useful for the previous analysis and annotation, peaks called in cellranger were merged, resulting in different length peaks. PyCisTopic needs annotated cell types to form pseudobulk profiles to call peaks on. I prefer this approach for pseudobulk differential analysis, so here the PyCisTopic peaks are applied for a new feature matrix to use for differential analysis.
```{r}
consensus_regions <- read.table("pycisTopic/outs/consensus_peak_calling/consensus_regions.bed", header = F, col.names = c("chr", "start", "end", "peak_id", "V1", "V2"))

consensus_regions$end - consensus_regions$start

consensus_regions
peaks.gr <- GenomicRanges::makeGRangesFromDataFrame(consensus_regions)

# I remove the unzipped cellranger outputs when I dont need them for a while. If the paths to the fragment.tsv changed I need to update them to get the fragment information (this is too big to save in the seurat object, only the path to the cellranger output is saved.)
new_frags <- list()
new_frags[[1]] <- Fragments(sarc_atac_integrated[["ATAC"]])[[1]] |>
  UpdatePath(new.path = "~/workbench/pool_1/fragments.tsv.gz")
new_frags[[2]] <- Fragments(sarc_atac_integrated[["ATAC"]])[[2]] |>
  UpdatePath(new.path = "~/workbench/pool_2_1/fragments.tsv.gz")
new_frags[[3]] <- Fragments(sarc_atac_integrated[["ATAC"]])[[3]] |>
  UpdatePath(new.path = "~/workbench/pool_2_2/fragments.tsv.gz")
new_frags[[4]] <- Fragments(sarc_atac_integrated[["ATAC"]])[[4]] |>
  UpdatePath(new.path = "~/workbench/pool_3_1/fragments.tsv.gz")
new_frags[[5]] <- Fragments(sarc_atac_integrated[["ATAC"]])[[5]] |>
  UpdatePath(new.path = "~/workbench/pool_3_2/fragments.tsv.gz")
# 
Fragments(sarc_atac_integrated[["ATAC"]]) <- NULL
Fragments(sarc_atac_integrated[["ATAC"]]) <- new_frags

Fragments(sarc_atac_pctopic[["ATAC"]]) <- NULL
Fragments(sarc_atac_pctopic[["ATAC"]]) <- new_frags


### build new celltype specific feature matrix (takes long)
counts <- FeatureMatrix(fragments = Fragments(sarc_atac_integrated[["ATAC"]]),
                        features = peaks.gr,
                        cells = Cells(sarc_atac_integrated))

chromatin_assay <- CreateChromatinAssay(counts = counts,
                                        fragments = Fragments(sarc_atac_integrated[["ATAC"]]))

ah <- AnnotationHub()

query(ah, "EnsDb.Hsapiens.v98")

ensdb_v98 <- ah[["AH75011"]]

annotations <- GetGRangesFromEnsDb(ensdb = ensdb_v98)

seqlevels(annotations) <- paste0("chr", seqlevels(annotations))
genome(annotations) <- "hg38"

Annotation(chromatin_assay) <- annotations

sarc_atac_pctopic <- CreateSeuratObject(chromatin_assay,
                                        assay = "ATAC",
                                        meta.data = sarc_atac_integrated@meta.data)
sarc_atac_pctopic
write_rds(sarc_atac_pctopic, "sarc_atac_pctopic.rds")
write_rds(sarc_atac_integrated, "sarc_atac_integrated.rds")
```

### Pseudobulk: Create DESeq datasets
```{r}
sample_info <- readxl::read_xlsx("../250804_Sarcoidosis_patient_metadata.xlsx") |> 
  mutate(pool = case_when(`scATAC pool` == "1.1" ~ "pool_1",
                          `scATAC pool` == "2.1" ~ "pool_2_1",
                          `scATAC pool` == "2.2" ~ "pool_2_2",
                          `scATAC pool` == "3.1" ~ "pool_3_1",
                          `scATAC pool` == "3.2" ~ "pool_3_2"))

pops <- sarc_atac_pctopic$cluster_names_pseudobulk |> unique()

pseudobulk <- sapply(pops, function(pop){
bulk <- sarc_atac_pctopic |> 
  subset(state != "CGD") |> 
  subset(cluster_names_pseudobulk == pop) |>
AggregateExpression(group.by = "SampleID", return.seurat = F, normalization.method = NULL)

bulk <- bulk$ATAC
colnames(bulk) <- str_replace_all(colnames(bulk), "-", "_")
colnames(bulk)[1:4] <- str_replace(colnames(bulk)[1:4], "g", "")

libsize <- bulk |>
  as.data.frame() |>
  summarise(across(everything(), ~ sum(.x))) |>
  t() |>
  as.data.frame() |>
  rename(libsize = V1) |> 
  rownames_to_column("SampleID")

sample_info_bulk <- sample_info |>
  mutate(state = factor(state, levels = c("healthy", "sarcoidosis"))) |> left_join(libsize)

bulk_anno <- data.frame(SampleID = colnames(bulk)) |> 
  left_join(sample_info_bulk)

dds <- DESeqDataSetFromMatrix(bulk, bulk_anno, ~state + pool + Sex)

peaks.keep <- rownames(dds)[!(str_detect(rownames(dds), "X-") | str_detect(rownames(dds), "Y-"))]
peaks.keep <- peaks.keep[str_detect(peaks.keep, "chr")]

dds <- dds[rownames(dds) %in% peaks.keep,]

detected <- dds |> 
  counts() |> 
  rowMeans()

detected <- detected[detected > 10] |> names()

dds <- dds[rownames(dds) %in% detected,]

dds <- DESeq(dds)
return(dds)
})

write_rds(pseudobulk, "output/pseudobulk.rds")
```

```{r}
all_fc <- lapply(names(pseudobulk), function(pop){
  pseudobulk[[pop]] |> 
    results(name = "state_sarcoidosis_vs_healthy") |> 
    as.data.frame() |> 
    rownames_to_column("peak") |> 
    mutate(ClosestFeature(sarc_atac_pctopic[["ATAC"]], peak),
           celltype = pop) 
}) |> bind_rows()

write_csv(all_fc, "output/all_fc.csv")
```

### Pseudobulk: ORA all pathways
```{r}
pops <- c("CD14_Mono", "CD4_memory", "CD4_naive", "CD8_memory", "NK")


hallmarks <- msigdbr(collection = "H")
go <- msigdbr(collection = "C5", subcollection = "GO:BP")
kegg <- msigdbr(collection = "C2", subcollection = "CP:KEGG_LEGACY")
reactome <- msigdbr(collection = "C2", subcollection = "CP:REACTOME")

allpws <- rbind(hallmarks, go)

all_fc <- read_csv("output/all_fc.csv")

ora_hm <- lapply(pops, function(pop){
res <- all_fc |> 
  subset(celltype == pop) |> 
  subset(!is.na(padj))

universe <- res$gene_name

ora_up <- enricher(res |> subset(padj < 0.05 & log2FoldChange > 0) |> pull(gene_name), 
         universe = universe, 
         TERM2GENE = allpws |> select(gs_id, gene_symbol),
         TERM2NAME = allpws |> select(gs_id, gs_name), 
         pAdjustMethod = "BH", 
         qvalueCutoff = 0.05)

ora_down <- enricher(res |> subset(padj < 0.05 & log2FoldChange < 0) |> pull(gene_name), 
         universe = universe, 
         TERM2GENE = allpws |> select(gs_id, gene_symbol),
         TERM2NAME = allpws |> select(gs_id, gs_name), 
         pAdjustMethod = "BH", 
         qvalueCutoff = 0.05)

ora_up <- ora_up@result |> 
  mutate(direction = "up")

ora_down <- ora_down@result |> 
  mutate(direction = "down")

ora_hm <- rbind(ora_up, ora_down) |> mutate(celltype = pop)

return(ora_hm)
}) |> bind_rows()

write_csv(ora_hm, "output/ora_all.csv")
```

## Tn5 insertion coverage:
#### write out barcodes per cell type and condition
```{r}
celltypes <- c("CD14_Mono", "CD4_naive", "CD4_memory", "CD8_memory", "NK")

celltype_barcodes <- sarc_atac_pctopic@meta.data |> 
  select(barcode, cluster_names_pseudobulk, pool, state) |> 
  mutate(celltype_pool = paste0(cluster_names_pseudobulk, "_", state, "_", pool),
         celltype = paste0(cluster_names_pseudobulk, "_", state),
         folder = case_when(pool == "pool_1" ~ "202403108_1_count",
                            pool == "pool_2_1" ~ "202403108a_1_count",
                            pool == "pool_2_2" ~ "202403108a_2_count",
                            pool == "pool_3_1" ~ "202403108b_3-1_count",
                            pool == "pool_3_2" ~ "202403108b_3-2_count")) |> 
  subset(cluster_names_pseudobulk %in% celltypes) |> 
  subset(state != "CGD") |> 
  rownames_to_column("barcode_pooled")

files <- celltype_barcodes$celltype_pool |> unique()
celltypes <- celltype_barcodes$celltype |> unique()

for (file in files){
  celltype_barcodes |> 
    subset(celltype == file) |> 
    pull(barcode) |> 
    write_lines(paste0("/home/titus.schluter@mydre.org/celltype_barcodes/", file, ".txt"))
}

celltype_barcodes |> 
  select(pool, folder) |> 
  unique()
```

### Write tsv fragments file for each celltype_state
```{bash, eval = F}
for pool in pool_1 pool_2_1 pool_2_2 pool_3_1 pool_3_2; do 
  zcat ${pool}/outs/fragments.tsv.gz | grep -F -f celltype_barcodes/${pop}_${pool}.txt > celltype_barcodes/${pop}_${pool}.tsv
done

for pop in CD8_memory_sarcoidosis CD14_Mono_sarcoidosis CD4_memory_sarcoidosis CD8_memory_healthy NK_sarcoidosis CD14_Mono_healthy CD4_naive_healthy NK_healthy CD4_naive_sarcoidosis CD4_memory_healthy; do
cat celltype_barcodes/${pop}_pool*.tsv > ${pop}_all.tsv
done

conda install -c bioconda ucsc-fetchchromsizes ucsc-bedgraphtobigwig

fetchChromSizes hg38 > hg38.chrom.sizes

for pop in CD8_memory_sarcoidosis CD14_Mono_sarcoidosis CD4_memory_sarcoidosis CD8_memory_healthy NK_sarcoidosis CD14_Mono_healthy CD4_naive_healthy NK_healthy CD4_naive_sarcoidosis CD4_memory_healthy; do
cat ${pop}_all.tsv | sed "s/chr1/chr01/g" | sed "s/chr2/chr02/g" | sed "s/chr3/chr03/g" | sed "s/chr4/chr04/g" | sed "s/chr5/chr05/g" | sed "s/chr6/chr06/g" | sed "s/chr7/chr07/g" | sed "s/chr8/chr08/g" | sed "s/chr9/chr09/g" > ${pop}_all_vorangestelltenullen.tsv
sort -V -k1,1 -k2,2n -k3,3n ${pop}_all_vorangestelltenullen.tsv > ${pop}_all.sorted.tsv
cat ${pop}_all.sorted.tsv | sed "s/chr01/chr1/g" | sed "s/chr02/chr2/g" | sed "s/chr03/chr3/g" | sed "s/chr04/chr4/g" | sed "s/chr05/chr5/g" | sed "s/chr06/chr6/g" | sed "s/chr07/chr7/g" | sed "s/chr08/chr8/g" | sed "s/chr09/chr9/g" > ${pop}.sorted.tsv
grep -v -E "^GL|^KI" ${pop}.sorted.tsv > ${pop}_nocontigs.sorted.tsv
done

for pop in CD8_memory_sarcoidosis CD14_Mono_sarcoidosis CD4_memory_sarcoidosis CD8_memory_healthy NK_sarcoidosis CD14_Mono_healthy CD4_naive_healthy NK_healthy CD4_naive_sarcoidosis CD4_memory_healthy; do
bedtools genomecov -i ${pop}_nocontigs.sorted.tsv -g hg38.chrom.sizes -bg > ${pop}.bedgraph
bedGraphToBigWig ${pop}.bedgraph hg38.chrom.sizes ${pop}.bw
done

for pop in CD8_memory_sarcoidosis CD14_Mono_sarcoidosis CD4_memory_sarcoidosis CD8_memory_healthy NK_sarcoidosis CD14_Mono_healthy CD4_naive_healthy NK_healthy CD4_naive_sarcoidosis CD4_memory_healthy; do
mv ${pop}.bw /mnt/data/2_Users/Titus_Schluter/${pop}.bw
done

for pop in CD4_naive CD4_memory CD8_memory CD14_Mono NK; do
rm ${pop}*
done

rm -r celltype_barcodes
rm hg38.chrom.sizes
```

## Tn5 insertion bed files
#### Write cell barcodes for each cell type and condition (state)
```{r}
## Create BigWig files to use in the UCSC genome browser
celltypes <- c("CD4_memory", "CD4_naive", "CD8_memory", "NK")

celltype_barcodes <- sarc_atac_pctopic@meta.data |> 
  subset(state %in% c("healthy", "sarcoidosis")) |> 
  select(barcode, cluster_names_pseudobulk, pool, state) |> 
  mutate(celltype_pool = paste0(cluster_names_pseudobulk, "_", state, "_", pool),
         celltype = paste0(cluster_names_pseudobulk, "_", state)) |> 
  subset(cluster_names_pseudobulk %in% celltypes) |> 
  rownames_to_column("barcode_pooled")

files <- celltype_barcodes$celltype_pool |> unique()
celltypes <- celltype_barcodes$celltype |> unique()

for (file in files){
  celltype_barcodes |> 
    subset(celltype_pool == file) |> 
    pull(barcode) |> 
    write_lines(paste0("Tn5_bed/", file, ".txt"))
}
```

# Generate Tn5 insertion bed files and generate bigwig of normalized coverage
```{bash, eval = F}
cd /mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/Tn5_bed

for pop in CD14_Mono_sarcoidosis CD14_Mono_healthy; do
for pool in pool_1 pool_2_1 pool_2_2 pool_3_1 pool_3_2; do 
  zcat ~/workbench/${pool}/fragments.tsv.gz | grep -F -f ${pop}_${pool}.txt > ${pop}_${pool}.tsv
done
done

for pop in CD14_Mono_sarcoidosis CD14_Mono_healthy; do
cat ${pop}_pool*.tsv > ${pop}_all.tsv
done


## python script from github.com/tacazares/scATAC_analysis/scripts
python fragmentsToTn5.py -i CD14_Mono_healthy_all.tsv -o /. -n CD14_Mono_healthy_Tn5
python fragmentsToTn5.py -i CD14_Mono_sarcoidosis_all.tsv -o /. -n CD14_Mono_sarcoidosis_Tn5

#conda install -c bioconda ucsc-fetchchromsizes ucsc-bedgraphtobigwig

fetchChromSizes hg38 > hg38.chrom.sizes

gunzip CD14_Mono_healthy_Tn5.bed.gz
gunzip CD14_Mono_sarcoidosis_Tn5.bed.gz

sort -k 1,1 CD14_Mono_healthy_Tn5.bed > CD14_Mono_healthy_Tn5_sorted.bed
sort -k 1,1 TCD14_Mono_sarcoidosis_Tn5.bed > CD14_Mono_sarcoidosis_Tn5_sorted.bed

bedtools genomecov -dz -i CD14_Mono_sarcoidosis_Tn5_sorted.bed -g hg38.chrom.sizes > CD14_Mono_sarcoidosis_Tn5.bedgraph
bedtools genomecov -dz -i CD14_Mono_healthy_Tn5_sorted.bed -g hg38.chrom.sizes > CD14_Mono_healthy_Tn5.bedgraph

N_healthy=$(wc -l < CD14_Mono_healthy_Tn5.bed)
N_sarcoidosis=$(wc -l < CD14_Mono_sarcoidosis_Tn5.bed)

scale_healthy=$(awk -v n=$N_healthy 'BEGIN{print 1000000.0/n}')
scale_sarcoidosis=$(awk -v n=$N_sarcoidosis 'BEGIN{print 1000000.0/n}')

awk 'BEGIN{OFS="\t"} {print $1, $2, $2+1, $3}' CD14_Mono_healthy_Tn5.bedgraph > CD14_Mono_healthy_Tn5_end.bedgraph
awk -v s=$scale_healthy 'BEGIN{OFS="\t"} {print $1, $2, $3, ($4+0.0)*s}' CD14_Mono_healthy_Tn5_end.bedgraph > CD14_Mono_healthy_Tn5_scaled.bedgraph
awk 'BEGIN{OFS="\t"} {print $1, $2, $2+1, $3}' CD14_Mono_sarcoidosis_Tn5.bedgraph > CD14_Mono_sarcoidosis_Tn5_end.bedgraph
awk -v s=$scale_sarcoidosis 'BEGIN{OFS="\t"} {print $1, $2, $3, ($4+0.0)*s}' CD14_Mono_sarcoidosis_Tn5_end.bedgraph > CD14_Mono_sarcoidosis_Tn5_scaled.bedgraph

bedGraphToBigWig CD14_Mono_healthy_Tn5_scaled.bedgraph ../hg38.chrom.sizes CD14_Mono_healthy_Tn5.bw -unc
bedGraphToBigWig CD14_Mono_sarcoidosis_Tn5_scaled.bedgraph ../hg38.chrom.sizes CD14_Mono_sarcoidosis_Tn5.bw -unc
```

same process for the other cell types in parallel
```{bash, eval = F}
cd /mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/Tn5_bed

for pop in CD4_naive_healthy CD4_naive_sarcoidosis CD4_memory_healthy CD4_memory_sarcoidosis CD8_memory_healthy CD8_memory_sarcoidosis NK_healthy NK_sarcoidosis; do
for pool in pool_1 pool_2_1 pool_2_2 pool_3_1 pool_3_2; do 
  zcat ~/workbench/${pool}/fragments.tsv.gz | grep -F -f ${pop}_${pool}.txt > ${pop}_${pool}.tsv
done
done

for pop in CD4_naive_healthy CD4_naive_sarcoidosis CD4_memory_healthy CD4_memory_sarcoidosis CD8_memory_healthy CD8_memory_sarcoidosis NK_healthy NK_sarcoidosis; do
cat ${pop}_pool*.tsv > ${pop}_all.tsv
done

parallel -j 4 "python ../fragmentsToTn5.py -i {1}_healthy_all.tsv -o . -n {1}_healthy_Tn5 \
python ../fragmentsToTn5.py -i {1}_sarcoidosis_all.tsv -o . -n {1}_sarcoidosis_Tn5" ::: CD4_naive CD4_memory CD8_memory NK

#conda install -c bioconda ucsc-fetchchromsizes ucsc-bedgraphtobigwig

fetchChromSizes hg38 > hg38.chrom.sizes

for pop in CD4_naive_healthy CD4_naive_sarcoidosis CD4_memory_healthy CD4_memory_sarcoidosis CD8_memory_healthy CD8_memory_sarcoidosis NK_healthy NK_sarcoidosis; do
gunzip ${pop}_Tn5.bed.gz
done

parallel -j 4 "sort -k 1,1 {1}_healthy_Tn5.bed > {1}_healthy_Tn5_sorted.bed" ::: CD4_naive CD4_memory CD8_memory NK

parallel -j 4 "sort -k 1,1 {1}_sarcoidosis_Tn5.bed > {1}_sarcoidosis_Tn5_sorted.bed" ::: CD4_naive CD4_memory CD8_memory NK

parallel -j 4 "bedtools genomecov -dz -i {1}_healthy_Tn5_sorted.bed -g hg38.chrom.sizes > {1}_healthy_Tn5.bedgraph" ::: CD4_naive CD4_memory CD8_memory NK

parallel -j 4 "bedtools genomecov -dz -i {1}_sarcoidosis_Tn5_sorted.bed -g hg38.chrom.sizes > {1}_sarcoidosis_Tn5.bedgraph" ::: CD4_naive CD4_memory CD8_memory NK


for pop in CD4_naive CD4_memory CD8_memory NK; do
N_healthy=$(wc -l < ${pop}_healthy_Tn5.bed)
N_sarcoidosis=$(wc -l < ${pop}_sarcoidosis_Tn5.bed)

scale_healthy=$(awk -v n=$N_healthy 'BEGIN{print 1000000.0/n}')
scale_sarcoidosis=$(awk -v n=$N_sarcoidosis 'BEGIN{print 1000000.0/n}')

awk 'BEGIN{OFS="\t"} {print $1, $2, $2+1, $3}' ${pop}_healthy_Tn5.bedgraph > ${pop}_healthy_Tn5_end.bedgraph
awk -v s=$scale_healthy 'BEGIN{OFS="\t"} {print $1, $2, $3, ($4+0.0)*s}' ${pop}_healthy_Tn5_end.bedgraph > ${pop}_healthy_Tn5_scaled.bedgraph
awk 'BEGIN{OFS="\t"} {print $1, $2, $2+1, $3}' ${pop}_sarcoidosis_Tn5.bedgraph > ${pop}_sarcoidosis_Tn5_end.bedgraph
awk -v s=$scale_sarcoidosis 'BEGIN{OFS="\t"} {print $1, $2, $3, ($4+0.0)*s}' ${pop}_sarcoidosis_Tn5_end.bedgraph > ${pop}_sarcoidosis_Tn5_scaled.bedgraph
done

parallel -j 4 "bedGraphToBigWig {1}_healthy_Tn5_scaled.bedgraph hg38.chrom.sizes {1}_healthy_Tn5.bw -unc" ::: CD4_naive CD4_memory CD8_memory NK
parallel -j 4 "bedGraphToBigWig {1}_sarcoidosis_Tn5_scaled.bedgraph hg38.chrom.sizes {1}_sarcoidosis_Tn5.bw -unc" ::: CD4_naive CD4_memory CD8_memory NK
```

## SCENICplus installation
```{bash, eval = f}
conda create --name scenicplus python=3.11.8

conda activate scenicplus
git clone https://github.com/aertslab/scenicplus
cd scenicplus
git checkout development
pip install .
```

## write DARs as input for SCENIC+
```{r}
pops = c("NK", "CD8_memory", "CD4_memory", "CD4_naive", "CD14_Mono")

for (pop in pops){
  all_fc |> 
    subset(celltype == pop) |> 
    subset(padj < 0.1) |> 
    arrange(padj) |> 
    subset(log2FoldChange > 0) %>%
    mutate(chromosome = word(peak, 1, sep = "-"),
           start = word(peak, 2, sep = "-"),
           end = word(peak, 3, sep = "-"),
           peak_id = paste0(celltype, "_up_peak_", 1:nrow(.))) |> 
    dplyr::select(chromosome, start, end, peak_id) |> 
    write_tsv(paste0("pycisTopic/outs/region_sets/DARs_celltype/", pop, "_up_peaks.bed"), col_names = F)
  
  all_fc |> 
    subset(celltype == pop) |> 
    subset(padj < 0.1) |> 
    arrange(padj) |> 
    subset(log2FoldChange < 0) %>%
    mutate(chromosome = word(peak, 1, sep = "-"),
           start = word(peak, 2, sep = "-"),
           end = word(peak, 3, sep = "-"),
           peak_id = paste0(celltype, "_down_peak_", 1:nrow(.))) |> 
    dplyr::select(chromosome, start, end, peak_id) |> 
    write_tsv(paste0("pycisTopic/outs/region_sets/DARs_celltype/", pop, "_down_peaks.bed"), col_names = F)
  
  ## Write all detected peaks per celltype (for HOMER):
  all_fc |> 
    subset(celltype == pop) |> 
    subset(!is.na(padj)) |> 
    mutate(chromosome = word(peak, 1, sep = "-"),
           start = word(peak, 2, sep = "-"),
           end = word(peak, 3, sep = "-")) |> 
    dplyr::select(chromosome, start, end) |> 
    write_tsv(paste0("HOMER/background_peaks/", pop, "_background_peaks.bed"), col_names = F)
}
```

## Find cell type marker regions for SCENIC+
```{r}
marker_regions <- FindAllMarkers(sarc_atac_pctopic, assay = "ATAC", min.pct = 0.1, group.by = "cluster_names_pseudobulk", only.pos = T)

selected <- marker_regions |> 
  group_by(cluster) |> 
  subset(p_val_adj < 1e-50) |> 
  slice_min(p_val_adj, n = 5000) |> 
  ungroup()

selected$cluster |> 
  unique() |> 
  as.vector() |> 
  lapply(function(pop){
    selected |> 
      subset(cluster == pop) %>% 
      mutate(chromosome = word(gene, 1, sep = "-"),
             start = word(gene, 2, sep = "-"),
             end = word(gene, 3, sep = "-"),
             peak_id = paste0(pop, "_marker_region", 1:nrow(.))) |> 
      dplyr::select(chromosome, start, end, peak_id) |> 
      write_tsv(paste0("pycisTopic/outs/region_sets/celltype_marker_regions/", pop, "_marker_regions.bed"), col_names = F)
  })
```

## create tss.bed for pycistopic
```{r}
tss <- read_tsv("pycisTopic/mart_export.txt")

tss <- tss |> 
  subset(`Transcript type` == "protein_coding") |> 
  rename(`# Chromosome` = `Chromosome/scaffold name`,
         End = `Transcription start site (TSS)`,
         Gene = `Gene name`,
         Transcript_type = `Transcript type`) |> 
  subset(`# Chromosome` %in% c("MT", "X", "Y", 1:22)) |> 
  mutate(`# Chromosome` = case_when(`# Chromosome` == "MT" ~ "chrM",
                                TRUE ~ paste0("chr", `# Chromosome`)),
         Start = End - 1,
         Strand = case_when(Strand == 1 ~ "+",
                            Strand == -1 ~ "-"),
         Score = ".") |> 
  select(`# Chromosome`, Start, End, Gene, Score, Strand, Transcript_type)
  
write_tsv(tss, "pycisTopic/tss.bed")


read_tsv("/home/titus.schluter@mydre.org/workbench/pool_1/outs/fragments.tsv.gz")
```

## SCENIC+ create custom cistarget database
```{bash}
# install create_cistarget_database
git clone https://github.com/aertslab/create_cisTarget_databases
cd create_cisTarget_databases
echo "create_cistarget_databases_dir='""${PWD}""'"
create_cistarget_databases_dir="/home/titus.schluter@mydre.org/create_cisTarget_databases"

conda create -n create_cistarget_databases \
    'python=3.10' \
    'numpy=1.21' \
    'pandas>=1.4.1' \
    'pyarrow>=7.0.0' \
    'numba>=0.55.1' \
    'python-flatbuffers'
    
conda activate create_cistarget_databases

# download cluster buster
cd "${CONDA_PREFIX}/bin"
wget https://resources.aertslab.org/cistarget/programs/cbust
# make the downloaded bin executable
chmod a+x cbust

# same with ucsc tools
# liftOver
wget http://hgdownload.soe.ucsc.edu/admin/exe/linux.x86_64/liftOver
wget http://hgdownload.soe.ucsc.edu/admin/exe/linux.x86_64/bigWigAverageOverBed

# Make downloaded binaries executable.
chmod a+x liftOver bigWigAverageOverBed

# download motif collection
mkdir -p aertslab_motif_colleciton
wget -O aertslab_motif_colleciton/v10nr_clust_public.zip https://resources.aertslab.org/cistarget/motif_collections/v10nr_clust_public/v10nr_clust_public.zip
cd aertslab_motif_colleciton
unzip -q v10nr_clust_public.zip
cd ..
# check available species motif databases
ls aertslab_motif_colleciton/v10nr_clust_public/snapshots/

# if not on the drive, download human genome fasta hg38.fa and human genome chrom.sizes
wget -O human_genome/hg38.fa.gz http://hgdownload.cse.ucsc.edu/goldenPath/hg38/bigZips/hg38.fa.gz
wget -O human_genome/hg38.fa.gz http://hgdownload.cse.ucsc.edu/goldenPath/hg38/bigZips/hg38.fa.gz
gunzip human_genome/hg38.fa.gz
wget -O human_genome/hg38.chrom.sizes http://hgdownload.cse.ucsc.edu/goldenPath/hg38/bigZips/hg38.chrom.sizes

# assemble a fasta from consensus peaks
REGION_BED="/mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/pycisTopic/outs/consensus_peak_calling/consensus_regions.bed"
GENOME_FASTA="/home/titus.schluter@mydre.org/human_genome/hg38.fa"
CHROMSIZES="/home/titus.schluter@mydre.org/human_genome/hg38.chrom.sizes"
DATABASE_PREFIX="sarcoidosis_atac"
SCRIPT_DIR="/home/titus.schluter@mydre.org/create_cisTarget_databases"

${SCRIPT_DIR}/create_fasta_with_padded_bg_from_bed.sh \
        ${GENOME_FASTA} \
        ${CHROMSIZES} \
        ${REGION_BED} \
        hg38.sarcoidosis_atac_consensus.fa \
        1000 \
        yes
        
# fetch all motifs from the collection
ls aertslab_motif_colleciton/v10nr_clust_public/singletons > motifs.txt

# create cistarget database (this may take long)
OUT_DIR=""${PWD}""
CBDIR="${OUT_DIR}/aertslab_motif_colleciton/v10nr_clust_public/singletons"
FASTA_FILE="${OUT_DIR}/hg38.sarcoidosis_atac_consensus.fa"
MOTIF_LIST="${OUT_DIR}/motifs.txt"

"${SCRIPT_DIR}/create_cistarget_motif_databases.py" \
    -f ${FASTA_FILE} \
    -M ${CBDIR} \
    -m ${MOTIF_LIST} \
    -o ${OUT_DIR}/${DATABASE_PREFIX} \
    --bgpadding 1000 \
    -t 20
```

## SCENIC+ config.yaml for the Snakemake pipeline
```{bash}
input_data:
  cisTopic_obj_fname: "/mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/pycisTopic/outs/cistopic_obj.pkl"
  GEX_anndata_fname: "/mnt/data/2_Users/Titus_Schluter/sarcoidosis_rna/scenicplus_adata.h5ad"
  region_set_folder: "/mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/pycisTopic/outs/region_sets/"
  ctx_db_fname: "/home/titus.schluter@mydre.org/workbench/scplus_sarcoidosis/sarcoidosis.regions_vs_motifs.rankings.feather"
  dem_db_fname: "/home/titus.schluter@mydre.org/workbench/scplus_sarcoidosis/sarcoidosis.regions_vs_motifs.scores.feather"
  path_to_motif_annotations: "/home/titus.schluter@mydre.org/aertslab_motif_colleciton/v10nr_clust_public/snapshots/motifs-v10-nr.hgnc-m0.00001-o0.0.tbl"

output_data:
  # output for prepare_GEX_ACC .h5mu
  combined_GEX_ACC_mudata: "/home/titus.schluter@mydre.org/workbench/outs/ACC_GEX.h5mu"
  # output for motif enrichment results .hdf5
  dem_result_fname: "/home/titus.schluter@mydre.org/workbench/outs/dem_results.hdf5"
  ctx_result_fname: "/home/titus.schluter@mydre.org/workbench/outs/ctx_results.hdf5"
  # output html for motif enrichment results .html
  output_fname_dem_html: "/home/titus.schluter@mydre.org/workbench/outs/dem_results.html"
  output_fname_ctx_html: "/home/titus.schluter@mydre.org/workbench/outs/ctx_results.html"
  # output for prepare_menr .h5ad
  cistromes_direct: "/home/titus.schluter@mydre.org/workbench/outs/cistromes_direct.h5ad"
  cistromes_extended: "/home/titus.schluter@mydre.org/workbench/outs/cistromes_extended.h5ad"
  # output tf names .txt
  tf_names: "/home/titus.schluter@mydre.org/workbench/outs/tf_names.txt"
  # output for download_genome_annotations .tsv
  genome_annotation: "/home/titus.schluter@mydre.org/workbench/outs/genome_annotation.tsv"
  chromsizes: "/home/titus.schluter@mydre.org/workbench/outs/chromsizes.tsv"
  # output for search_space .tsb
  search_space: "/home/titus.schluter@mydre.org/workbench/outs/search_space.tsv"
  # output tf_to_gene .tsv
  tf_to_gene_adjacencies: "/home/titus.schluter@mydre.org/workbench/outs/tf_to_gene_adj.tsv"
  # output region_to_gene .tsv
  region_to_gene_adjacencies: "/home/titus.schluter@mydre.org/workbench/outs/region_to_gene_adj.tsv"
  # output eGRN .tsv
  eRegulons_direct: "/home/titus.schluter@mydre.org/workbench/outs/eRegulon_direct.tsv"
  eRegulons_extended: "/home/titus.schluter@mydre.org/workbench/outs/eRegulons_extended.tsv"
  # output AUCell .h5mu
  AUCell_direct: "/home/titus.schluter@mydre.org/workbench/outs/AUCell_direct.h5mu"
  AUCell_extended: "/home/titus.schluter@mydre.org/workbench/outs/AUCell_extended.h5mu"
  # output scplus mudata .h5mu
  scplus_mdata: "/home/titus.schluter@mydre.org/workbench/outs/scplusmdata.h5mu"

params_general:
  temp_dir: "/home/titus.schluter@mydre.org/workbench/tmp"
  n_cpu: 55
  seed: 666

params_data_preparation:
  # Params for prepare_GEX_ACC
  bc_transform_func: "\"lambda x: f'{x}'\""
  is_multiome: False
  key_to_group_by: "cluster_names_pseudobulk"
  nr_cells_per_metacells: 10
  # Params for prepare_menr
  direct_annotation: "Direct_annot"
  extended_annotation: "Orthology_annot"
  # Params for download_genome_annotations
  species: "hsapiens"
  biomart_host: "http://www.ensembl.org"
  # Params for search_space
  search_space_upstream: "1000 150000"
  search_space_downstream: "1000 150000"
  search_space_extend_tss: "10 10"

params_motif_enrichment:
  species: "homo_sapiens"
  annotation_version: "v10nr_clust"
  motif_similarity_fdr: 0.001
  orthologous_identity_threshold: 0.0
  annotations_to_use: "Direct_annot Orthology_annot"
  fraction_overlap_w_dem_database: 0.4
  dem_max_bg_regions: 500
  dem_balance_number_of_promoters: True
  dem_promoter_space: 1_000
  dem_adj_pval_thr: 0.05
  dem_log2fc_thr: 1.0
  dem_mean_fg_thr: 0.0
  dem_motif_hit_thr: 3.0
  fraction_overlap_w_ctx_database: 0.4
  ctx_auc_threshold: 0.005
  ctx_nes_threshold: 3.0
  ctx_rank_threshold: 0.05




params_inference:
  # Params for tf_to_gene
  tf_to_gene_importance_method: "GBM"
  # Params regions_to_gene
  region_to_gene_importance_method: "GBM"
  region_to_gene_correlation_method: "SR"
  # Params for eGRN inference
  order_regions_to_genes_by: "importance"
  order_TFs_to_genes_by: "importance"
  gsea_n_perm: 1000
  quantile_thresholds_region_to_gene: "0.85 0.90 0.95"
  top_n_regionTogenes_per_gene: "5 10 15"
  top_n_regionTogenes_per_region: ""
  min_regions_per_gene: 0
  rho_threshold: 0.05
  min_target_genes: 10
```

## Generate DORC scores from SCENIC+ results
```{r}
library(FigR)

eRegulons <- read_tsv("outs_scplus/eRegulon_direct.tsv")

eRegulons

DORC_df <- eRegulons |> 
  select(Region, Gene, rho_R2G) |> 
  subset(rho_R2G > 0.50) |> 
  unique()

DORC_df |> 
  select(Gene, Region) |> 
  unique() |> 
  count(Gene) |> 
  arrange(-n) %>%
  mutate(rank = 1:nrow(.),
         label = Gene == "CEBPB") |> 
  ggplot(aes(x = rank, y = n)) +
  geom_point(color = "steelblue") +
  theme_classic() +
  force_panelsizes(rows = unit(2, "in"),
                   cols = unit(2, "in")) +
  geom_hline(yintercept = 9.5) +
  geom_text_repel(aes(label = ifelse(label, Gene, "")), hjust = -1, max_overlaps = 15)
  
DORC_genes <- DORC_df |> 
  select(Gene, Region) |> 
  unique() |> 
  count(Gene) |> 
  subset(n >= 10) |> 
  pull(Gene)

library(FigR)
library(SummarizedExperiment)
library(GenomicRanges)

ATAC.se <- SummarizedExperiment(
  assay = list(counts = sarc_atac_pctopic[["ATAC"]]$counts),
  rowRanges = granges(sarc_atac_pctopic),
  colData = sarc_atac_pctopic[[]]
)

DORC_df_filtered <- DORC_df |> 
  subset(Gene %in% DORC_genes) |> 
  mutate(Peak = match(Region |> str_replace(":", "-"), rownames(ATAC.se)))

DORC_scores <- FigR::getDORCScores(ATAC.se,
                                   dorcTab = DORC_df_filtered, 
                                   geneList = DORC_genes,
                                   normalizeATACmat = TRUE, nCores = 50)

sarc_atac_pctopic[["DORC"]] <- CreateAssayObject(counts = DORC_scores)
```

# HOMER
```{bash, eval = F}
# Homer bugs if you have @ in your username, which is the case for all default MyDRE accounts, so you have to make a new account and ssh into it to install and run homer:
sudo adduser titus
sudo usermod -a -G titus

# ssh into new user:
exit
ssh titus@10.210.167.148

# Homer setup. Note this is where the @ symbol issues for our DRE usernames appear. 
mkdir ~/HOMER
cd ~/HOMER
wget homer.ucsd.edu/homer/configureHomer.pl
perl ~/HOMER/configureHomer.pl -install
perl ~/HOMER/configureHomer.pl -install hg38
```

### Run HOMER on SCENIC+ DARs
```{bash}
# Running homer. This only runs on my special user without the @ symbol in the name (nicholas). Takes about 30min (run up and down in parallel as limited to 1 core).
PATH=$PATH:~/HOMER/bin

parallel "~/HOMER/bin/findMotifsGenome.pl /mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/pycisTopic/outs/region_sets/DARs_celltype/{1}_up_peaks.bed hg38 /mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/HOMER/{1}_up_peaks -size 200 -bg /mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/HOMER/background_peaks/{1}_background_peaks.bed" ::: CD4_naive CD4_memory CD8_memory CD14_Mono NK

parallel "~/HOMER/bin/findMotifsGenome.pl /mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/pycisTopic/outs/region_sets/DARs_celltype/{1}_down_peaks.bed hg38 /mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/HOMER/{1}_down_peaks -size 200 -bg /mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/HOMER/background_peaks/{1}_background_peaks.bed" ::: CD4_naive CD4_memory CD8_memory CD14_Mono NK
```

# fragmentsToTn5.py (github.com/tacazares/scATAC_analysis)
```{python}
#!/usr/bin/python3

### Convert scATAC-seq fragments to a Tn5 bed file ###

# Inputs:

# input_fragments: Path to input fragment tsv file
# output directory: Path to directory to write results
# name: Output file name
# split: Split fragment ends into different files

# Outputs:

# A bed file of Tn5 sites that have been inferred from the scATAC-seq fragments file.
# If split is True: Then output two bed files, one for each end of the fragment.

######################################################

###-Imports-###
import pandas as pd
import argparse
import os

###-Arguments-###
parser = argparse.ArgumentParser()

parser.add_argument("-i", "--input_fragments",
                dest='FRAGMENTS', 
                help="Input fragments file",
                required=True)

parser.add_argument("-o", "--output_dir",
                dest='OUT_DIR', 
                help="Output directory",
                default="./output",
                required=False)

parser.add_argument("-n", "--name",
                dest='NAME', 
                help="Name to use for output file",
                default="Tn5_IS",
                required=False)

parser.add_argument("-split", "--split",
                action='store_true',
                dest='SPLIT', 
                help="Split fragment ends into seperate bed files",
                default="False",
                required=False)

args = parser.parse_args()

###-Functions-###
def convert_fragments_to_tn5_bed(fragments_tsv, chroms, split=False):
    """Convert scATAC fragments file to Tn5 insertion sites bed

    Args:
        fragments_tsv (str): Path to the scATAC-seq fragments file

    Returns:
        dataframe: Tn5 insertions sites in a BED formatted pandas dataframe
    """
    col_types = {
                 "chr": "category", 
                 "start": "int32", 
                 "stop": "int32", 
                 "barcode": "category", 
                 "support": "int32"
                }
    
    # Import fragments tsv as a dataframe
    df = pd.read_table(fragments_tsv,
                       header=None,
                       names=["chr", "start", "stop", "barcode", "support"],
                       dtype=col_types, 
                       low_memory=False)
    
    df = df[df["chr"].isin(chroms)]
    
    # subset starts
    df_starts = df[["chr", "start", "barcode"]].copy()
    
    # subset stops
    df_ends = df[["chr", "stop", "barcode"]].copy()
    
    # Add a 1 to create 1 bp intervals representing the cut site
    df_starts["stop"] = df_starts["start"].copy() + 1
    
    # Subtract a 1 bp interval to represent the cut site. Reference miralidlab wiki page for scATAC-seq analysis
    df_ends["start"] = df_ends["stop"].copy() - 1
        
    if split:
        # If split is True, return two dataframes, one for each end of the fragment    
        return df_starts[["chr", "start", "stop", "barcode"]], df_ends[["chr", "start", "stop", "barcode"]]
    
    else:
        # Concatenate both dataframes        
        df_cat = pd.concat([df_starts, df_ends])

        return df_cat[["chr", "start", "stop", "barcode"]]

###-Constants-###
ALL_CHRS = [
            "chr1", "chr2", "chr3", "chr4", "chr5", "chr6",
            "chr7", "chr8", "chr9", "chr10", "chr11", "chr12",
            "chr13", "chr14", "chr15", "chr16", "chr17", "chr18",
            "chr19", "chr20", "chr21", "chr22", "chrX", "chrY"
            ]

###-Workflow-###
if args.SPLIT == True:
    # Create file names
    output_file_start = os.path.join(args.OUT_DIR, args.NAME + "_start.bed.gz")
    output_file_stop = os.path.join(args.OUT_DIR, args.NAME + "_stop.bed.gz")

    print("Splitting Fragment ends into individual files")
    
    start_interval_df, stop_interval_df = convert_fragments_to_tn5_bed(args.FRAGMENTS, ALL_CHRS, args.SPLIT)
    
    start_interval_df.to_csv(output_file_start, sep="\t", header=False, index=False)
    
    stop_interval_df.to_csv(output_file_stop, sep="\t", header=False, index=False)

else:
    # Create file names
    output_file = os.path.join(args.OUT_DIR, args.NAME + ".bed.gz")
    print("Splitting Fragment ends into a new BED file: " + output_file)

    tn5_bed = convert_fragments_to_tn5_bed(args.FRAGMENTS, ALL_CHRS)

    tn5_bed.to_csv(output_file, sep="\t", header=False, index=False)
```