---
title: "Sarcoidosis_snATAC"
format: html
editor: source
---

Install packages and check downloaded files

# scAVENGERs genotype demux
#### Install scAVENGERS and run it for genotype demux of each pool
```{bash, eval = F}
wget https://github.com/kaistcbfg/scAVENGERS/archive/refs/tags/v1.0.0.tar.gz

tar -xvzf v1.0.0.tar.gz

conda env create -f scAVENGERS-1.0.0/envs/environment.yml

wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_47/GRCh38.primary_assembly.genome.fa.gz

gunzip GRCh38.primary_assembly.genome.fa.gz

conda activate scavengers
mkdir ~/outdir
conda install samtools

samtools faidx GRCh38.primary_assembly.genome.fa

# Modified ~/scAVENGERS-1.0.0/scripts/get_programs.sh. Added PATH=$PATH:/home/titus.schluter@mydre.org/miniconda3/envs/scavengers/bin
# Modified ~/scAVENGERS-1.0.0/snakefile. Changed vcftools in line 49 to bcftools
```

#### Example of config.yaml used for demultiplexing
```{bash, eval = F}
# number of threads to use
# to make this valid, you must set the number of cores when running snakemake pipeline.
# e.g. -j 10
THREADS: 24

# input and output data path
DATA:
  # path to directory for saving output files
  out_dir:
    /home/titus.schluter@mydre.org/outdir
  # path to genome sequence file
  genome:
    /home/titus.schluter@mydre.org/GRCh38.primary_assembly.genome.fa
  # path to alignment file
  alignment:
    /home/titus.schluter@mydre.org/202403108b_3-2_count/outs/possorted_bam.bam
  # path to text file containing line-seperated list of barcodes
  barcode:
    /home/titus.schluter@mydre.org/202403108b_3-2_count/outs/filtered_peak_bc_matrix/barcodes.tsv

# cluster.py settings
CLUSTER:
  # number of genotypes in multiplexed sample
  n_genotypes: 4
  # ploidy of the organism of interest
  ploidy: 2
  # error rate correction parameter of probability model
  err_rate: 0.001
  # difference of likelihood to stop the iteration
  stop_criterion: 0.1
  # maximum number of iteration
  max_iter: 1000

# variant caller settings
VARIANT_CALLER:
  # variant caller to use. Choose "freebayes" or "strelka."
  caller: "freebayes"
  # minimum quality score of variant, used only if the caller is "freebayes"
  min_qual: 0
  # whether to use low-GQX variants only, used only if the caller is "strelka"
  lowgqx: False

# vartrix settings
VARTRIX:
  # minimum mapping quality of read to use in vartrix
  mapq: 30

# troublet settings
TROUBLET:
  # doublet prior
  doublet_prior: 0.5
  # doublet posterior threshold
  doublet_threshold: 0.9
  # singlet posterior threshold
  singlet_threshold: 0.9
```

#### running scavengers
```{bash, eval = F}
~/scAVENGERS-1.0.0/scAVENGERS pipeline --configfile ~/scAVENGERS-1.0.0/config.yaml -j $THREADS
```

#### checking relatedness of genotyped individuals and scAVENGERS clusters

after demultiplexing, we compare the cluster_genotypes.vcf for relatedness with genotyped individuals

```{bash, eval = F}
# Preparing conda environment for plink and bcftools.
# This has already been done
conda deactivate
conda config --add channels bioconda
conda config --add channels conda-forge
conda config --set channel_priority strict
conda create -n plink
conda activate plink
conda install plink
conda install bcftools
```

```{bash}
# Activate conda environment.
conda activate plink

export WORKING_DIR=/mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac

# Preparing genotype files for each pool.
for Pool in pool_1 pool_2_1 pool_2_2 pool_3_1 pool_3_2
do
# Creating genotype directory.
rm -r $WORKING_DIR/scAVENGERS/$Pool/genotypes
mkdir $WORKING_DIR/scAVENGERS/$Pool/genotypes

# Updating sample IDs in cluster genotypes file.
touch $WORKING_DIR/scAVENGERS/$Pool/genotypes/new_sample_ids.txt
echo cluster0 >> $WORKING_DIR/scAVENGERS/$Pool/genotypes/new_sample_ids.txt
echo cluster1 >> $WORKING_DIR/scAVENGERS/$Pool/genotypes/new_sample_ids.txt
echo cluster2 >> $WORKING_DIR/scAVENGERS/$Pool/genotypes/new_sample_ids.txt
echo cluster3 >> $WORKING_DIR/scAVENGERS/$Pool/genotypes/new_sample_ids.txt

bcftools reheader --samples $WORKING_DIR/scAVENGERS/$Pool/genotypes/new_sample_ids.txt -o $WORKING_DIR/scAVENGERS/$Pool/genotypes/cluster_genotypes2.vcf $WORKING_DIR/scAVENGERS/$Pool/cluster_genotypes.vcf

# Converting to plink format.
plink --vcf $WORKING_DIR/scAVENGERS/$Pool/genotypes/cluster_genotypes2.vcf --allow-extra-chr --set-missing-var-ids @_#_\$1_\$2 --chr 1-25 --make-bed --out $WORKING_DIR/scAVENGERS/$Pool/genotypes/cluster_genotypes2
done
```

```{r}
# Identifying overlapping variants between actual genotypes and cluster genotypes.
for(Pool in c("pool_1", "pool_2_1", "pool_2_2", "pool_3_1", "pool_3_2")) {
  # Reading in cluster genotypes file.
  cluster_geno_bim <- read_delim(paste0("scAVENGERS/", Pool, "/genotypes/cluster_genotypes2.bim"), delim = "\t", col_names = F)
  
  # Reading in lifted actual genotypes.
  actual_geno_bim2 <- read_delim("/mnt/data/2_Users/Nicholas_Sumpter/Genotypes_April_2025/For_scAVENGERS/actual_geno_lifted.bim", delim = "\t", col_names = F)
  
  # Finding overlap of SNPs using only chromosome, position, a1, and a2 (not bothering with flipped alleles or strand switches for now).
  overlap_snps <- actual_geno_bim2 |> 
    inner_join(cluster_geno_bim, by = c("X1", "X4", "X5", "X6")) |> 
    mutate(CHR_BP_A1_A2 = paste(X1, X4, X5, X6, sep = "_"))
  
  # Excluding duplicated sites.
  tmp <- overlap_snps |> 
    filter(duplicated(CHR_BP_A1_A2) | duplicated(CHR_BP_A1_A2, fromLast = T))
  
  overlap_snps_dedup <- overlap_snps |> 
    filter(!CHR_BP_A1_A2 %in% tmp$CHR_BP_A1_A2)
  
  # Filtering to keep overlapped SNPs.
  cluster_geno_bim_overlap <- cluster_geno_bim |> 
    mutate(CHR_BP_A1_A2 = paste(X1, X4, X5, X6, sep = "_")) |> 
    filter(CHR_BP_A1_A2 %in% overlap_snps_dedup$CHR_BP_A1_A2)
  
  actual_geno_bim2_overlap <- actual_geno_bim2 |> 
    mutate(CHR_BP_A1_A2 = paste(X1, X4, X5, X6, sep = "_")) |> 
    filter(CHR_BP_A1_A2 %in% overlap_snps_dedup$CHR_BP_A1_A2)
  
  write_delim(cluster_geno_bim_overlap |> select(X2), paste0("scAVENGERS/", Pool, "/genotypes/cluster_geno_overlap_snps.txt"), delim = "\t", col_names = F)
  
  write_delim(actual_geno_bim2_overlap |> select(X2), paste0("scAVENGERS/", Pool, "/genotypes/actual_geno2_overlap_snps.txt"), delim = "\t", col_names = F)
  
  rm(actual_geno_bim2, cluster_geno_bim, overlap_snps, overlap_snps_dedup, tmp)
}

rm(Pool, cluster_geno_bim_overlap, actual_geno_bim2_overlap)
```

```{bash, eval = F}
# Extracting overlap SNPs from each file.
for Pool in pool_1 pool_2_1 pool_2_2 pool_3_1 pool_3_2
do
plink --bfile /mnt/data/2_Users/Nicholas_Sumpter/Genotypes_April_2025/For_scAVENGERS/actual_geno_lifted --extract $WORKING_DIR/scAVENGERS/$Pool/genotypes/actual_geno2_overlap_snps.txt --make-bed --out $WORKING_DIR/scAVENGERS/$Pool/genotypes/actual_geno_overlap

plink --bfile $WORKING_DIR/scAVENGERS/$Pool/genotypes/cluster_genotypes2 --extract $WORKING_DIR/scAVENGERS/$Pool/genotypes/cluster_geno_overlap_snps.txt --make-bed --out $WORKING_DIR/scAVENGERS/$Pool/genotypes/cluster_geno_overlap
done
```

```{r}
# Creating bim files with corrected names for merging and creating fam files with singular family IDs.
for(Pool in c("pool_1", "pool_2_1", "pool_2_2", "pool_3_1", "pool_3_2")) {
  # Creating bim files with corrected names for merging.
  actual_geno_bim2_overlap <- read_delim(paste0("scAVENGERS/", Pool, "/genotypes/actual_geno_overlap.bim"), delim = "\t", col_names = F) |> 
    mutate(CHR_BP_A1_A2 = paste(X1, X4, X5, X6, sep = "_"))
  
  cluster_geno_bim_overlap <- read_delim(paste0("scAVENGERS/", Pool, "/genotypes/cluster_geno_overlap.bim"), delim = "\t", col_names = F) |> 
    mutate(CHR_BP_A1_A2 = paste(X1, X4, X5, X6, sep = "_"))
  
  write_delim(actual_geno_bim2_overlap |> select(X1, CHR_BP_A1_A2, X3:X6), paste0("scAVENGERS/", Pool, "/genotypes/actual_geno_formerge.bim"), delim = "\t", col_names = F)
  
  write_delim(cluster_geno_bim_overlap |> select(X1, CHR_BP_A1_A2, X3:X6), paste0("scAVENGERS/", Pool, "/genotypes/cluster_geno_formerge.bim"), delim = "\t", col_names = F)
  
  # Creating fam files with 1 family ID.
  actual_geno_fam <- read_delim(paste0("scAVENGERS/", Pool, "/genotypes/actual_geno_overlap.fam"), col_names = F) |> 
    mutate(X1 = 0)
  
  cluster_geno_fam <- read_delim(paste0("scAVENGERS/", Pool, "/genotypes/cluster_geno_overlap.fam"), col_names = F) |> 
    mutate(X1 = 0)
  
  write_delim(actual_geno_fam, paste0("scAVENGERS/", Pool, "/genotypes/actual_geno_formerge.fam"), delim = " ", col_names = F)
  
  write_delim(cluster_geno_fam, paste0("scAVENGERS/", Pool, "/genotypes/cluster_geno_formerge.fam"), delim = " ", col_names = F)
  
  rm(cluster_geno_fam, actual_geno_fam, actual_geno_bim2_overlap, cluster_geno_bim_overlap)
}

rm(Pool)
```

```{bash, eval = F}
# Merging files and calculating relatedness.
for Pool in pool_1 pool_2_1 pool_2_2 pool_3_1 pool_3_2
do
# Copying bed files.
cp $WORKING_DIR/scAVENGERS/$Pool/genotypes/actual_geno_overlap.bed $WORKING_DIR/scAVENGERS/$Pool/genotypes/actual_geno_formerge.bed
cp $WORKING_DIR/scAVENGERS/$Pool/genotypes/cluster_geno_overlap.bed $WORKING_DIR/scAVENGERS/$Pool/genotypes/cluster_geno_formerge.bed

# Merging files.
plink --bfile $WORKING_DIR/scAVENGERS/$Pool/genotypes/cluster_geno_formerge --bmerge $WORKING_DIR/scAVENGERS/$Pool/genotypes/actual_geno_formerge --make-bed --out $WORKING_DIR/scAVENGERS/$Pool/genotypes/merge_geno

# Preparing independent SNPs for genome relatedness.
plink --bfile $WORKING_DIR/scAVENGERS/$Pool/genotypes/merge_geno --indep 50 5 2 --out $WORKING_DIR/scAVENGERS/$Pool/genotypes/merge_geno

plink --bfile $WORKING_DIR/scAVENGERS/$Pool/genotypes/merge_geno --extract $WORKING_DIR/scAVENGERS/$Pool/genotypes/merge_geno.prune.in --make-bed --out $WORKING_DIR/scAVENGERS/$Pool/genotypes/merge_geno2

# Running plink genome tool.
plink --bfile $WORKING_DIR/scAVENGERS/$Pool/genotypes/merge_geno2 --genome --out $WORKING_DIR/scAVENGERS/$Pool/genotypes/merge_geno_plinkgenome
done
```

#### Matching scavengers clusters to genotypes
```{r}
# Loading in resulting plink genome files.
plink_rel <- read_table("scAVENGERS/pool_1/genotypes/merge_geno_plinkgenome.genome") |> 
  mutate(Pool = "") |> 
  slice(0)

for(Pool in c("pool_1", "pool_2_1", "pool_2_2", "pool_3_1", "pool_3_2")) {
  tmp <- read_table(paste0("scAVENGERS/", Pool, "/genotypes/merge_geno_plinkgenome.genome")) |> 
    mutate(Pool = Pool)
  
  plink_rel <- rbind(plink_rel, tmp)
}

plink_rel <- plink_rel |> 
  #filter(str_detect(IID1, "Pat|cluster") & str_detect(IID2, "Pat|cluster")) |> 
  mutate(PI_HAT = as.numeric(PI_HAT)) |> 
  subset(PI_HAT > 0.3) |> 
  select(Pool, IID1, IID2, PI_HAT, RATIO)

rm(tmp, Pool)
```

# R setup
```{r}
library(ggpmisc)
library(msigdbr)
library(ggrepel)
library(ggh4x)
library(harmony)
library(JASPAR2020)
library(TFBSTools)
library(DESeq2)
library(motifmatchr)
library(BSgenome.Hsapiens.UCSC.hg38)
library(clusterProfiler)
library(ggpubr)
library(pheatmap)
library(ggbeeswarm)
library(AnnotationHub)
library(GenomicRanges)
library(Seurat)
library(Signac)
library(org.Hs.eg.db)
library(tidyverse)

setwd("/mnt/data/2_Users/Titus_Schluter/sarcoidosis_atac/")
```


```{r}
#sarc_atac_integrated <- readRDS("sarc_atac_integrated.rds")
#sarc_atac_pctopic <- readRDS("sarc_atac_pctopic.rds")
```

```{r}
DefaultAssay(sarc_atac_integrated)

sarc_atac_integrated$test <- factor(sarc_atac_integrated$cluster_names_pseudobulk, levels = c("CD14_Mono",
                                                                                              "CD16_Mono",
                                                                                              "DC",
                                                                                              "pDC",
                                                                                              "CD4_naive",
                                                                                              "CD4_memory",
                                                                                              "Treg",
                                                                                              "CD8_naive",
                                                                                              "CD8_memory",
                                                                                              "MAIT",
                                                                                              "NKT_like",
                                                                                              "NK",
                                                                                              "early_NK",
                                                                                              "B_naive",
                                                                                              "B_memory") |> rev())

Idents(sarc_atac_integrated) <- "test"

DotPlot(sarc_atac_integrated |> subset(cluster_names_pseudobulk != "excl"),
        features = c("CD33",
                     "ITGAM",
                     "CD14",
                     "ITGAX",
                     "FCGR3A",
                     "CD1C",
                     "IL3RA",
                     "CD3E",
                     "CD4",
                     "CCR7",
                     "CD69",
                     "CD27",
                     "IL7R",
                     "CCR4",
                     "IL2RA",
                     "FOXP3",
                     "CD8A",
                     "CCL5",
                     "KLRB1",
                     "NCAM1",
                     "CD19",
                     "MS4A1",
                     "IGHD",
                     "IGHM"),
        cluster.idents = F) +
  scale_color_gradient2(low = "#6fa8dc", mid = "grey90", high = "#e06666") +
  scale_radius(limits = c(0, 100)) +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        panel.grid = element_blank()) +
  force_panelsizes(rows = unit(3.5 , "in"),
                   cols = unit(5 * 22/24, "in"))

ggsave("celltype_annotation_atac.pdf", height = 8, width = 16)
```

For Signac I need to merge multiple objects and unify peaks. Peaks were initially called by cellranger and are different for each experiment. In this chunk, first peaks are unified and then based on the fragments files new count matrices are generated. From count matrices and fragments files separate seurat objects are generated which all contain the same features. Then, seurat objects can be merged.

# Find Consensus peaks and combine data
```{r}
gr <- GRanges(seqnames = "chr1", ranges = IRanges(start = c(20, 70, 300), end = c(120, 200, 400)))
gr

experiments <- list.files("raw_data/unzipped/")

peaks <- lapply(experiments, function(exp){
  read.table(paste0("raw_data/unzipped/", exp, "/outs/peaks.bed"),
             col.names = c("chr", "start", "end"))
})

peaks.gr <- lapply(peaks, function(exp){
  makeGRangesFromDataFrame(exp)
})

combined.peaks <- GenomicRanges::reduce(x = c(peaks.gr[[1]],
                               peaks.gr[[2]],
                               peaks.gr[[3]],
                               peaks.gr[[4]],
                               peaks.gr[[5]]))

peakwidths <- width(combined.peaks)
peakwidths |> hist()

width(peaks.gr[[1]]) |> hist()

metadata <- lapply(experiments, function(exp){
  md <- read.table(file = paste0("raw_data/unzipped/", exp, "/outs/singlecell.csv"),
             stringsAsFactors = F,
             sep = ",",
             header = T,
             row.names = 1)[-1, ] |> 
    mutate(exp.name = exp,
           pool = case_when(exp.name == "202403108_1_count" ~ "pool_1",
                            exp.name == "202403108a_1_count" ~ "pool_2_1",
                            exp.name == "202403108a_2_count" ~ "pool_2_2",
                            exp.name == "202403108b_3-1_count" ~ "pool_3_1",
                            exp.name == "202403108b_3-2_count" ~ "pool_3_2"))
  md$cellID <- rownames(md)
  return(md)
})

fragment_files <- lapply(experiments, function(exp){
  cells <- read.table(paste0("raw_data/unzipped/", exp, "/outs/filtered_peak_bc_matrix/barcodes.tsv.gz"), header = F)
  cells <- cells$V1
  CreateFragmentObject(path = paste0("raw_data/unzipped/", exp, "/outs/fragments.tsv.gz"),
                       cells = cells)
})

counts <- list()

for (i in 1:length(experiments)){
  cells <- read.table(paste0("raw_data/unzipped/", experiments[i], "/outs/filtered_peak_bc_matrix/barcodes.tsv.gz"), header = F)
  cells <- cells$V1
  counts[i] <- FeatureMatrix(fragments = fragment_files[i],
                features = combined.peaks,
                cells = cells)
}

#write_rds(counts, "signac_counts.rds")

chrom_assays <- list()

for (i in 1:length(experiments)){
  chrom_assays[i] <- CreateChromatinAssay(counts[[i]],
                                          fragments = fragment_files[[i]])
}

seurat_objects <- list()

for (i in 1:length(experiments)){
  seurat_objects[[i]] <- CreateSeuratObject(chrom_assays[[i]],
                                          assay = "ATAC",
                                          meta.data = metadata[[i]])
}

sarc_atac <- merge(x = seurat_objects[[1]], y = seurat_objects[2:length(seurat_objects)])
sarc_atac
#write_rds(sarc_atac, "sarc_atac.rds")
```

Filtering and annotating features
```{r}
# Filtering peaks on chromosome scaffolds and non standard chromosomes
peaks.keep <- seqnames(granges(sarc_atac)) %in% standardChromosomes(granges(sarc_atac))
sarc_atac <- sarc_atac[as.vector(peaks.keep),]

ah <- AnnotationHub()

query(ah, "EnsDb.Hsapiens.v98")

ensdb_v98 <- ah[["AH75011"]]

annotations <- GetGRangesFromEnsDb(ensdb = ensdb_v98)

seqlevels(annotations) <- paste0("chr", seqlevels(annotations))
genome(annotations) <- "hg38"

Annotation(sarc_atac) <- annotations
```

```{r}
sarc_atac <- NucleosomeSignal(sarc_atac)
sarc_atac <- TSSEnrichment(sarc_atac)

sarc_atac$pct_reads_in_peaks <- sarc_atac$peak_region_fragments / sarc_atac$passed_filters

sarc_atac$blacklist_ratio <- FractionCountsInRegion(sarc_atac,
                                                    assay = "ATAC",
                                                    regions = blacklist_hg38_unified)

sarc_atac$logCount_ATAC <- log10(sarc_atac$nCount_ATAC)
sarc_atac$logFeature_ATAC <- log10(sarc_atac$nFeature_ATAC)
```

### assign cell barcodes to demultiplexed clusters and add metadata
```{r}
pools = c("pool_1", "pool_2_1", "pool_2_2", "pool_3_1", "pool_3_2")

assignments <- lapply(pools, function(pool){
  read.table(paste0(pool, "_outdir/clusters.final.tsv"), header = T) |> 
    mutate(pool = pool)
}) |> 
  bind_rows() |> 
  select(barcode, status, assignment, pool) |> 
  mutate(pool_barcode = paste0(pool, "_", barcode),
         cluster = case_when(status == "singlet" ~ paste0("cluster", assignment),
                             status == "doublet" ~ "doublet"),
         pool_cluster = case_when(status == "singlet" ~ paste0(pool, "_", cluster),
                                  status == "doublet" ~ "doublet"))
```

### QC and filtering
```{r}
DensityScatter(sarc_atac, x = "nCount_ATAC", "TSS.enrichment", log_x = T, quantiles = T)

VlnPlot(sarc_atac, features = c("nCount_ATAC"), pt.size = 0, group.by = "pool") +
  geom_hline(aes(yintercept = 5000))
ggsave("output/qc_nCount_ATAC.pdf", height = 4, width = 7)
VlnPlot(sarc_atac, features = c("TSS.enrichment"), pt.size = 0, group.by = "pool") +
  geom_hline(aes(yintercept = 4))
ggsave("output/qc_TSS_enrichment.pdf", height = 4, width = 7)
VlnPlot(sarc_atac, features = c("blacklist_ratio"), pt.size = 0, group.by = "pool") +
  geom_hline(aes(yintercept = 0.005))
ggsave("output/qc_blacklist_ratio.pdf", height = 4, width = 7)
VlnPlot(sarc_atac, features = c("nucleosome_signal"), pt.size = 0, group.by = "pool") +
  geom_hline(aes(yintercept = 3))
ggsave("output/qc_nucleosome_signal.pdf", height = 4, width = 7)
VlnPlot(sarc_atac, features = c("pct_reads_in_peaks"), pt.size = 0, group.by = "pool") +
  geom_hline(aes(yintercept = 0.4))
ggsave("output/qc_pct_reads_in_peaks.pdf", height = 4, width = 7)

FragmentHistogram(sarc_atac_filtered, group.by = "pool")
ggsave("output/fragment_sizes.pdf", height =4, width =7)

# write_rds(sarc_atac, "sarc_atac_filtered.rds")

sarc_atac_filtered <- sarc_atac |> 
  subset(nCount_ATAC > 5000) |> 
  subset(TSS.enrichment > 4) |>
  subset(blacklist_ratio < 0.005) |> 
  subset(nucleosome_signal < 3) |> 
  subset(pct_reads_in_peaks > 0.4) |> 
  subset(status != "doublet") |> 
  subset(status != "unassigned")

ncol(sarc_atac)
ncol(sarc_atac_filtered)
```

### linear dimensional reduction
```{r}
sarc_atac_filtered <- RunTFIDF(sarc_atac_filtered)
sarc_atac_filtered <- FindTopFeatures(sarc_atac_filtered)
sarc_atac_filtered <- RunSVD(sarc_atac_filtered)

# in SVD dimensional reduction the first LSI often correlates strongly with sequencing depth as a technical artifact and should be excluded from downstream nonlinear dimensional reduction.
DepthCor(sarc_atac_filtered)
ggsave("output/depth_correlation_lsi.pdf", height = 4, width = 7)
ElbowPlot(sarc_atac_filtered, reduction = "lsi")
ggsave("output/elbow_plot_lsi.pdf", height = 4, width = 7)
```

### UMAP and harmony integration
```{r}
sarc_atac_filtered <- RunUMAP(sarc_atac_filtered, reduction = "lsi", dims = 2:20)
DimPlot(sarc_atac_filtered, group.by = "Patient")
DimPlot(sarc_atac_filtered, group.by = "pool")

sarc_atac_filtered[["ATAC_v5"]] <- sarc_atac_filtered[["ATAC"]]
DefaultAssay(sarc_atac_filtered) <- "ATAC_v5"

sarc_atac_filtered[["ATAC_v5"]] <- split(sarc_atac_filtered[["ATAC_v5"]], f = sarc_atac_filtered$Patient)

sarc_atac_integrated <- IntegrateLayers(sarc_atac_filtered,
                                        method = HarmonyIntegration,
                                        orig.reduction = "lsi",
                                        new.reduction = "harmony")

sarc_atac_integrated <- RunUMAP(sarc_atac_integrated,
                              reduction = "harmony",
                              dims = 2:20,
                              reduction.name = "umap.harmony")

#write_rds(sarc_atac_filtered, "sarc_atac_integrated.rds")
```

### Clustering
```{r}
DefaultAssay(sarc_atac_integrated) <- "ATAC"
sarc_atac_integrated <- FindNeighbors(sarc_atac_integrated, reduction = "harmony", dims = 2:20)
sarc_atac_integrated <- FindClusters(sarc_atac_integrated, resolution = 1, cluster.name = "clusters.1.harmony")


DimPlot(sarc_atac_integrated, reduction = "umap.harmony", group.by = "clusters.1.harmony")
#write_rds(sarc_atac_integrated, "sarc_atac_integrated.rds")
DefaultAssay(sarc_atac_integrated) <- "ATAC_v5"
```

### Gene activity inference
```{r}
DefaultAssay(sarc_atac_integrated) <- "ATAC"
gene.activity <- GeneActivity(sarc_atac_integrated)

sarc_atac_integrated[["iRNA"]] <- CreateAssayObject(gene.activity)

sarc_atac_integrated <- NormalizeData(sarc_atac_integrated, assay = "iRNA")
# write_rds(sarc_atac_integrated, "sarc_atac_integrated.rds")
```

### transfer labels from RNA data
```{r}
sarc_rna <- readRDS("/mnt/data/2_Users/Titus_Schluter/sarcoidosis_rna/sarcoidosis_integrated.rds")

DefaultAssay(sarc_atac_integrated) <- "iRNA"

FeaturePlot(sarc_atac_integrated, features = "CD3D", reduction = "umap.harmony")

transfer.anchors <- FindTransferAnchors(reference = sarc_rna,
                                        query = sarc_atac_integrated,
                                        reduction = "cca")

predicted.labels <- TransferData(anchorset = transfer.anchors,
                                 refdata = sarc_rna$cluster_names_pseudobulk,
                                 weight.reduction = sarc_atac_integrated[["harmony"]],
                                 dims = 2:20)
```

```{r}
metadata <- predicted.labels |>
select(predicted.id) |> 
  cbind(sarc_atac_integrated@meta.data)

sarc_atac_integrated@meta.data <- metadata

DimPlot(object = sarc_atac_integrated, 
              reduction = "umap.harmony", group.by = "predicted.id",
              label = TRUE,
              label.size = 3,
              raster = F) + NoLegend()
```

### label clusters based on label transfer
```{r}
# Labeling our clusters based on Azimuth labels.
Idents(sarc_atac_integrated) <- sarc_atac_integrated$clusters.1.harmony

tmp <- tibble(Var1 = "",
              Freq = 0,
              cluster_id = 0) |> 
  slice(-1)

for(i in unique(sarc_atac_integrated$clusters.1.harmony)) {
  tmp1 <- table(sarc_atac_integrated@meta.data |> 
                  filter(clusters.1.harmony == i) |>
                  pull(predicted.id)) |>
    as.data.frame() |>
    mutate(cluster_id = i)
  tmp <- rbind(tmp, tmp1)
}

tmp <- tmp |> 
  select(Cluster_ID = cluster_id, predicted.id = Var1, Freq) |> 
  mutate(Cluster_ID = as.integer(Cluster_ID)) |> 
  arrange(Cluster_ID, desc(Freq))

tmp2 <- tmp |> 
  group_by(Cluster_ID) |> 
  summarize(N = sum(Freq))

tmp3 <- tmp |> 
  left_join(tmp2) |> 
  mutate(Percent = (Freq / N) * 100) |> 
  filter(Percent > 20)

tmp4 <- tmp3 |> 
  select(Cluster_ID, predicted.id) |> 
  group_by(Cluster_ID) |> 
  summarise(cluster_names1 = paste(predicted.id, collapse = "/")) |> 
  mutate(cluster_names1 = make.unique(cluster_names1, sep = " "),
         cluster_names1 = factor(cluster_names1),
         Cluster_ID = factor(Cluster_ID))

sarc_atac_integrated@meta.data <- sarc_atac_integrated@meta.data |> 
  rownames_to_column("cell") |> 
  #select(-c(cluster_names1.x)) |> 
  left_join(tmp4, by = c("clusters.1.harmony" = "Cluster_ID")) |> 
  as.data.frame() |> 
  column_to_rownames("cell")

# sarc_atac_integrated@meta.data <- sarc_atac_integrated@meta.data |>
#   select(-c(cluster_names1.x, cluster_names1.y, cluster_names1.y.y))
```

```{r}
Idents(sarc_atac_integrated) <- sarc_atac_integrated$clusters.1.harmony
p1 <- DimPlot(object = sarc_atac_integrated, 
              reduction = "umap.harmony", 
              label = TRUE,
              label.size = 3,
              raster = F) + NoLegend()

p3 <- DimPlot(object = sarc_atac_integrated, 
              reduction = "umap.harmony", 
              label = TRUE,
              label.size = 3, group.by = "predicted.id",
              raster = F) + NoLegend()

p4 <- DimPlot(object = sarc_atac_integrated, 
              reduction = "umap.harmony", 
              label = TRUE,
              label.size = 3, group.by = "cluster_names_pseudobulk",
              raster = F) + NoLegend()

p2 <- DotPlot(sarc_atac_integrated, features =  c("CD3E", "CD4", "CD8A", "CCR7", "CD69", "IL2RA", "FOXP3", "KLRB1", "NCAM1", "FCGR3A", "CD19", "MS4A1", "IGHD", "IGHM", "CD27", "CD33", "ITGAM", "CD14", "ITGAX", "CD1C", "IL3RA", "CD34", "IL7R", "MKI67", "CCR4", "CCL5", "ITGA2B"), cluster.idents = T, ) +
  scale_color_gradient2(low = "blue", mid = "grey90", high = "red") +
  scale_radius(limits = c(0, 100)) +
  RotatedAxis() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank())

p1 + p3 + p4

#ggsave("output/clustering.pdf", height = 8, width = 24)
```

### subcluster CD4 T memory cells to get to the Tregs

Tregs may be more difficult to find in ATAC data. Based on the label transfers, clustering the atac data at a resolution of 1 did not split Tregs from other T cell clusters. Because of this, the CD4 T cells were subclustered.
```{r}
CD4_memory <- sarc_atac_integrated |> subset(cluster_names1 == "CD4_memory/Treg" |
                                             cluster_names1 == "CD4_memory" |
                                             cluster_names1 == "CD4_memory 1")

DefaultAssay(CD4_memory) <- "ATAC"
CD4_memory <- FindNeighbors(CD4_memory, reduction = "harmony", dims = 2:20)
CD4_memory <- FindClusters(CD4_memory, resolution = 1, cluster.name = "CD4.subclusters.1.harmony")

DimPlot(CD4_memory, reduction = "umap.harmony", group.by = "CD4.subclusters.1.harmony") +
DimPlot(CD4_memory, reduction = "umap.harmony", group.by = "predicted.id")

# Labeling our clusters based on transferred labels.
Idents(CD4_memory) <- CD4_memory$CD4.subclusters.1.harmony

tmp <- tibble(Var1 = "",
              Freq = 0,
              cluster_id = 0) |> 
  slice(-1)

for(i in unique(CD4_memory$CD4.subclusters.1.harmony)) {
  tmp1 <- table(CD4_memory@meta.data |> 
                  filter(CD4.subclusters.1.harmony == i) |>
                  pull(predicted.id)) |>
    as.data.frame() |>
    mutate(cluster_id = i)
  tmp <- rbind(tmp, tmp1)
}

tmp <- tmp |> 
  select(Cluster_ID = cluster_id, predicted.id = Var1, Freq) |> 
  mutate(Cluster_ID = as.integer(Cluster_ID)) |> 
  arrange(Cluster_ID, desc(Freq))

tmp2 <- tmp |> 
  group_by(Cluster_ID) |> 
  summarize(N = sum(Freq))

tmp3 <- tmp |> 
  left_join(tmp2) |> 
  mutate(Percent = (Freq / N) * 100) |> 
  filter(Percent > 20)

tmp4 <- tmp3 |> 
  select(Cluster_ID, predicted.id) |> 
  group_by(Cluster_ID) |> 
  summarise(cluster_names1 = paste(predicted.id, collapse = "/")) |> 
  mutate(CD4_subcluster_names1 = make.unique(cluster_names1, sep = " "),
         CD4_subcluster_names1 = factor(cluster_names1),
         Cluster_ID = factor(Cluster_ID))

CD4_memory@meta.data <- CD4_memory@meta.data |> 
  rownames_to_column("cell") |> 
  #select(-c(cluster_names1.x)) |> 
  left_join(tmp4, by = c("CD4.subclusters.1.harmony" = "Cluster_ID")) |> 
  as.data.frame() |> 
  column_to_rownames("cell")

add <- CD4_memory@meta.data |> 
  rownames_to_column("cell") |> 
  select(cell, CD4_subcluster_names1) |> 
  mutate(CD4_subcluster_names1 = case_when(CD4_subcluster_names1 == "CD4_memory/CD4_naive" ~ "CD4_memory",
                                           TRUE ~ CD4_subcluster_names1))

sarc_atac_integrated@meta.data <- sarc_atac_integrated@meta.data |> 
  rownames_to_column("cell") |> 
  left_join(add) |> 
  column_to_rownames("cell")


# sarc_atac_integrated@meta.data <- sarc_atac_integrated@meta.data |>
#   select(-c(cluster_names1.x, cluster_names1.y, cluster_names1.y.y))
```

### assign manual pseudobulk cluster names
clustering and T cell subclustering annotations are merged, these were the annotations used for pseudobulk
```{r}
sarc_atac_integrated@meta.data <- sarc_atac_integrated@meta.data |> 
  mutate(cluster_names_pseudobulk = case_when(cluster_names1 == "CD14_Mono 1" ~ "CD14_Mono",
                                              cluster_names1 == "CD14_Mono 2" ~ "CD14_Mono",
                                              cluster_names1 == "excl/CD14_Mono" ~ "excl",
                                              cluster_names1 == "CD8_memory 1" ~ "CD8_memory",
                                              cluster_names1 == "CD4_memory" ~ CD4_subcluster_names1,
                                              cluster_names1 == "CD4_memory/Treg" ~ CD4_subcluster_names1,
                                              cluster_names1 == "CD4_memory 1" ~ CD4_subcluster_names1,
                                              cluster_names1 == "CD4_naive 1" ~ "CD4_naive",
                                              cluster_names1 == "early_NK/NK" ~ "early_NK",
                                              TRUE ~ cluster_names1
                                              ))
```

## prepare pycisTopic ATAC preprocessing
```{r}
metadata <- sarc_atac_integrated[[]]

metadata |> 
  mutate(VSN_sample_id = pool) |> 
  write_tsv("pycisTopic/cell_data.tsv")
```

This was followed by pycistopic analysis (jupyter notebook)